UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2018-12-06 ms/main.tex
edited abstract

\begin{abstract}
\adpcomment{Changed the abstract.}
Genetic structure in cohorts results from technical, sampling and demographic variation, necessitating visualization as an early step in genomic analyses. Existing approaches are stymied by the massive size of modern cohorts and unbalanced sizes of populations of many disparate ancestries and struggle to place individuals of admixed ancestries. They also struggle to reconcile local genetic relationships in a global context, and the push for larger and more diverse genomic data will only exacerbate these issues. We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly reveal cryptic population structure, the flow between layers of diversity, and phenotype heterogeneity between populations. Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP intuitively clusters closely related individuals while placing them in a global continuum of genetic variation. We uncover previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population, opening new lines of investigation for demographic research and statistical genetics. We also demonstrate that in a massive and diverse cohort, UMAP outperforms state-of-the-art dimension reduction. Given its small computational cost and promising results, PCA-UMAP should become a standard approach to exploratory analysis of population-scale genetic datasets.
\end{abstract}

>>>>

\begin{abstract}
\adpcomment{Changed the abstract.}
Genetic structure in cohorts results from technical, sampling and demographic variation, necessitating visualization as an early step in genomic analyses. Existing approaches are stymied by the massive size of modern cohorts and unbalanced sizes of populations of many disparate ancestries. They also struggle to reconcile local genetic relationships with the global context. The push for larger and more diverse genomic data will only exacerbate these issues. We investigate an approach to dimension reduction of genomic data that combines principal components analysis (PCA) with uniform manifold approximation and projection (UMAP) to succinctly reveal cryptic population structure, the flow between layers of diversity\sgcomment{What does that even mean?}, and phenotype heterogeneity between populations. Using data from large-scale genomic datasets, we demonstrate that PCA-UMAP intuitively clusters closely related individuals while placing them in a global continuum of genetic variation. We uncover previously overlooked subpopulations within the American Hispanic population and fine-scale relationships between geography, genotypes, and phenotypes in the UK population, opening new lines of investigation for demographic research and statistical genetics. We also demonstrate that in a massive and diverse cohort, UMAP outperforms state-of-the-art dimension reduction methods for both global convergence and preservation of qualitative features of the data. Given its small computational cost and promising results, we think that UMAP should be run in tandem with PCA as a standard initial analysis of population-scale genetic datasets.
\end{abstract}
